---
title: "NBA Career Outcomes Prediction"
format: html
---

```{r setup, message=FALSE, warning=FALSE}
library(hoopR)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
library(nnet)
library(ranger)
library(MASS)
library(rstan)
library(bayesplot)
library(caret)
library(knitr)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```

## Project Overview

Given a player's first few seasons in the NBA, predict their career outcome tier:

- **Out of League**: washed out quickly, minimal career MPG
- **Role Player**: bench contributor, 10–20 MPG
- **Starter**: regular starter, 20+ MPG sustained
- **All-Star**: ≥1 All-Star selection

We will use multinomial logistic regression and random forest models with features derived from players' first 1–3 seasons.

---

## Data Collection

Pull league-wide per-season stats using `hoopR::nba_leaguedashplayerstats()` for seasons 2000–2015. Cache to disk to avoid re-scraping.

```{r load-season-stats, message=FALSE, warning=FALSE}
seasons <- 2000:2015
cache_file <- "data/season_stats.rds"

if (!file.exists(cache_file)) {
  dir.create("data", showWarnings = FALSE)
  season_stats_raw <- map_dfr(seasons, function(yr) {
    Sys.sleep(0.6)  # be polite to the API
    tryCatch(
      nba_leaguedashplayerstats(season = year_to_season(yr)) |>
        pluck("LeagueDashPlayerStats") |>
        mutate(season = yr),
      error = function(e) NULL
    )
  })
  saveRDS(season_stats_raw, cache_file)
} else {
  season_stats_raw <- readRDS(cache_file)
}
```

```{r clean-stats}
season_stats <- season_stats_raw |>
  transmute(
    player_id  = as.integer(PLAYER_ID),
    player     = PLAYER_NAME,
    season     = season,
    gp         = as.numeric(GP),
    min_pg     = as.numeric(MIN),
    ppg        = as.numeric(PTS),
    rpg        = as.numeric(REB),
    apg        = as.numeric(AST),
    spg        = as.numeric(STL),
    bpg        = as.numeric(BLK),
    fg_pct     = as.numeric(FG_PCT),
    fg3_pct    = as.numeric(FG3_PCT),
    ft_pct     = as.numeric(FT_PCT),
    age        = as.numeric(AGE)
  ) |>
  filter(gp >= 10)  # drop injury-shortened stints
```

## Career Outcome Labels

Label each player by their peak career tier. Pull each player's full career stats once, then classify.

```{r career-labels, message=FALSE, warning=FALSE}
all_star_cache <- "data/allstar_playerids.rds"

if (!file.exists(all_star_cache)) {
  # All-Star rosters are available via nba_allstarballotpredictor or
  # we approximate using seasons where a player averaged 20+ PPG and 3+ ASG appearances.
  # Use nba_leaguedashplayerstats filtered to All-Star seasons (2000-2015) as a proxy:
  # Players who appeared in the top scoring tier consistently are labeled All-Star.
  # For a cleaner label, pull each player's career accolades via nba_playercareerstats.
  player_ids <- unique(season_stats$player_id)

  career_raw <- map(player_ids, function(pid) {
    Sys.sleep(0.4)
    tryCatch({
      res <- nba_playercareerstats(player_id = pid)
      list(
        totals  = res |> pluck("CareerTotalsRegularSeason"),
        seasons = res |> pluck("SeasonTotalsRegularSeason")
      )
    }, error = function(e) NULL)
  })

  saveRDS(career_raw, all_star_cache)
} else {
  career_raw <- readRDS(all_star_cache)
}

# Cache was saved as a flat list of CareerTotalsRegularSeason data frames
# with a lowercase player_id column already attached
career_totals <- career_raw |>
  compact() |>
  bind_rows() |>
  transmute(
    player_id  = as.integer(player_id),
    career_gp  = as.numeric(GP),
    career_min = as.numeric(MIN),
    career_ppg = as.numeric(PTS) / as.numeric(GP),
    career_mpg = as.numeric(MIN) / as.numeric(GP),
    # Approximate seasons from observed season stats (no season-by-season data in cache)
    career_seasons = career_gp / 55
  )

# All-Star approximation: players with career PPG > 18 and 10+ seasons → All-Star
# Starter: career MPG >= 28 | PPG >= 14
# Role Player: career MPG 15–28 and career_seasons >= 4
# Out of League: everything else (short careers, low minutes)
player_labels <- career_totals |>
  mutate(
    outcome = case_when(
      career_ppg >= 18 & career_seasons >= 8             ~ "All-Star",
      career_mpg >= 28 | (career_ppg >= 14 & career_seasons >= 6) ~ "Starter",
      career_mpg >= 15 & career_seasons >= 4              ~ "Role Player",
      TRUE                                                 ~ "Out of League"
    ),
    outcome = factor(outcome,
                     levels = c("Out of League", "Role Player", "Starter", "All-Star"))
  ) |>
  dplyr::select(player_id, outcome, career_ppg, career_mpg, career_seasons)
```

```{r join-draft}
# Pull draft info for each player
draft_cache <- "data/draft_info.rds"

if (!file.exists(draft_cache)) {
  draft_raw <- map(unique(season_stats$player_id), function(pid) {
    Sys.sleep(0.003)
    tryCatch(
      nba_commonplayerinfo(player_id = pid) |>
        pluck("CommonPlayerInfo") |>
        mutate(player_id = pid),
      error = function(e) NULL
    )
  })
  saveRDS(draft_raw, draft_cache)
} else {
  draft_raw <- readRDS(draft_cache)
}

draft_info <- draft_raw |>
  compact() |>
  bind_rows() |>
  transmute(
    player_id  = as.integer(player_id),
    draft_pick = suppressWarnings(as.numeric(DRAFT_NUMBER))
  ) |>
  filter(!is.na(draft_pick))
```

```{r build-model-data}
# Identify each player's debut season, then keep their first 3 seasons
debut_season <- season_stats |>
  group_by(player_id) |>
  summarise(debut = min(season), .groups = "drop")

first3 <- season_stats |>
  left_join(debut_season, by = "player_id") |>
  filter(season <= debut + 2) |>
  group_by(player_id) |>
  summarise(
    across(c(ppg, rpg, apg, spg, bpg, fg_pct, fg3_pct, ft_pct, min_pg),
           mean, na.rm = TRUE, .names = "avg_{.col}"),
    n_seasons = n(),
    .groups = "drop"
  )

model_data <- first3 |>
  inner_join(player_labels, by = "player_id") |>
  inner_join(draft_info,   by = "player_id") |>
  filter(n_seasons >= 1)

glimpse(model_data)
```

---

## Exploratory Visualizations

### Visualization 1: Draft Position by Career Outcome

```{r viz1-draft, fig.width=8, fig.height=5}
model_data |>
  filter(draft_pick <= 60) |>
  ggplot(aes(x = outcome, y = draft_pick, fill = outcome)) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.8) +
  geom_jitter(width = 0.12, size = 1.5, alpha = 0.4, color = "gray30") +
  scale_y_reverse(breaks = c(1, 10, 20, 30, 45, 60)) +
  scale_fill_manual(values = c(
    "Out of League" = "#d73027",
    "Role Player"   = "#fc8d59",
    "Starter"       = "#4575b4",
    "All-Star"      = "#1a9850"
  )) +
  labs(
    title    = "Draft Position by Career Outcome",
    subtitle = "Lower pick number = higher draft position  |  2000–2015 draft classes",
    x = "Career Outcome", y = "Draft Pick Number"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

### Visualization 2: PPG in First 3 Seasons by Career Outcome

```{r viz2-ppg, fig.width=8, fig.height=5}
model_data |>
  ggplot(aes(x = avg_ppg, fill = outcome, color = outcome)) +
  geom_density(alpha = 0.35, linewidth = 0.8) +
  scale_fill_manual(values = c(
    "Out of League" = "#d73027",
    "Role Player"   = "#fc8d59",
    "Starter"       = "#4575b4",
    "All-Star"      = "#1a9850"
  )) +
  scale_color_manual(values = c(
    "Out of League" = "#d73027",
    "Role Player"   = "#fc8d59",
    "Starter"       = "#4575b4",
    "All-Star"      = "#1a9850"
  )) +
  labs(
    title    = "PPG Distribution (First 3 Seasons) by Career Outcome",
    subtitle = "Average over each player's first 1–3 seasons",
    x = "Points Per Game", y = "Density",
    fill = "Outcome", color = "Outcome"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

### Visualization 3: Career Scoring Trajectory by Outcome Tier

```{r viz3-trajectories, fig.width=9, fig.height=5, message=FALSE, warning=FALSE}
# For a random sample of players per tier, show their season PPG over first 3 years
set.seed(42)
sampled_ids <- model_data |>
  group_by(outcome) |>
  slice_sample(n = 10) |>
  pull(player_id)

traj_data <- season_stats |>
  inner_join(debut_season, by = "player_id") |>
  filter(player_id %in% sampled_ids, season <= debut + 2) |>
  mutate(year_in_league = season - debut + 1) |>
  inner_join(dplyr::select(player_labels, player_id, outcome), by = "player_id")

ggplot(traj_data, aes(x = year_in_league, y = ppg,
                       group = player_id, color = outcome)) +
  geom_line(alpha = 0.4, linewidth = 0.8) +
  geom_point(alpha = 0.6, size = 2) +
  stat_summary(aes(group = outcome), fun = mean,
               geom = "line", linewidth = 2, linetype = "dashed") +
  scale_color_manual(values = c(
    "Out of League" = "#d73027",
    "Role Player"   = "#fc8d59",
    "Starter"       = "#4575b4",
    "All-Star"      = "#1a9850"
  )) +
  scale_x_continuous(breaks = 1:3, labels = c("Year 1", "Year 2", "Year 3")) +
  labs(
    title    = "Scoring Trajectory in First 3 Seasons by Career Outcome",
    subtitle = "Thin = individual players (sample of 10/tier)  |  Dashed = tier average",
    x = "Season in League", y = "Points Per Game",
    color = "Outcome"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "right")
```

---

## Modeling

### Train/Test Split

```{r split}
set.seed(42)
model_data_clean <- model_data |>
  dplyr::select(outcome, draft_pick, avg_ppg, avg_rpg, avg_apg, avg_spg, avg_bpg,
         avg_fg_pct, avg_fg3_pct, avg_ft_pct, avg_min_pg, n_seasons) |>
  na.omit()

train_idx  <- createDataPartition(model_data_clean$outcome, p = 0.8, list = FALSE)
train_data <- model_data_clean[train_idx, ]
test_data  <- model_data_clean[-train_idx, ]

cat("Train:", nrow(train_data), " | Test:", nrow(test_data), "\n")
table(train_data$outcome)
```

---

### Model 1: Multinomial Logistic Regression

```{r multinom-fit, message=FALSE, warning=FALSE}
mlr_cache <- "models/mlr_fit.rds"
if (!file.exists(mlr_cache)) {
  set.seed(42)
  mlr_fit <- multinom(
    outcome ~ draft_pick + avg_ppg + avg_rpg + avg_apg + avg_spg + avg_bpg +
      avg_fg_pct + avg_fg3_pct + avg_ft_pct + avg_min_pg + n_seasons,
    data  = train_data,
    trace = FALSE,
    maxit = 500
  )
  saveRDS(mlr_fit, mlr_cache)
} else {
  mlr_fit <- readRDS(mlr_cache)
}
summary(mlr_fit)
```

```{r multinom-bootstrap}
boot_cache <- "models/mlr_boot.rds"
if (!file.exists(boot_cache)) {
  # Bootstrap 95% CIs on coefficients (500 resamples)
  set.seed(42)
  B <- 500
  boot_coefs <- replicate(B, {
    idx  <- sample(nrow(train_data), replace = TRUE)
    fit  <- multinom(
      outcome ~ draft_pick + avg_ppg + avg_rpg + avg_apg + avg_spg + avg_bpg +
        avg_fg_pct + avg_fg3_pct + avg_ft_pct + avg_min_pg + n_seasons,
      data  = train_data[idx, ],
      trace = FALSE, maxit = 500
    )
    coef(fit)
  }, simplify = FALSE)
  saveRDS(boot_coefs, boot_cache)
} else {
  boot_coefs <- readRDS(boot_cache)
}

# Combine into array and compute CIs
boot_array <- simplify2array(boot_coefs)   # [classes, predictors, B]
ci_lower <- apply(boot_array, c(1, 2), quantile, 0.025)
ci_upper <- apply(boot_array, c(1, 2), quantile, 0.975)

# Tidy coefficient table for display
coef_df <- as.data.frame(coef(mlr_fit)) |>
  tibble::rownames_to_column("outcome_class") |>
  tidyr::pivot_longer(-outcome_class, names_to = "predictor", values_to = "estimate") |>
  mutate(
    lower = as.vector(t(ci_lower))[match(paste(outcome_class, predictor),
                                         outer(rownames(ci_lower),
                                               colnames(ci_lower), paste))],
    upper = as.vector(t(ci_upper))[match(paste(outcome_class, predictor),
                                         outer(rownames(ci_upper),
                                               colnames(ci_upper), paste))]
  )
```

```{r multinom-eval}
mlr_pred <- predict(mlr_fit, newdata = test_data)
mlr_cm   <- confusionMatrix(mlr_pred, test_data$outcome)
cat("Multinomial LR — Test Accuracy:", round(mlr_cm$overall["Accuracy"], 3), "\n")
print(mlr_cm$table)
```

```{r multinom-coef-plot, fig.width=10, fig.height=5}
# Show coefficients for key predictors across outcome classes
key_preds <- c("avg_ppg", "avg_min_pg", "draft_pick", "avg_rpg", "avg_apg")

coef(mlr_fit) |>
  as.data.frame() |>
  tibble::rownames_to_column("outcome_class") |>
  tidyr::pivot_longer(-outcome_class, names_to = "predictor", values_to = "estimate") |>
  filter(predictor %in% key_preds) |>
  ggplot(aes(x = predictor, y = estimate, fill = outcome_class)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(
    "Role Player" = "#fc8d59", "Starter" = "#4575b4", "All-Star" = "#1a9850"
  )) +
  labs(title = "Multinomial LR Coefficients (vs. Out of League baseline)",
       x = NULL, y = "Coefficient", fill = "Class") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

---

### Model 2: Random Forest

```{r rf-fit}
rf_cache <- "models/rf_fit.rds"
if (!file.exists(rf_cache)) {
  set.seed(42)
  p <- ncol(train_data) - 1   # number of predictors

  # Simple CV to choose mtry
  mtry_candidates <- c(floor(sqrt(p)), floor(p / 3))
  cv_oob <- sapply(mtry_candidates, function(m) {
    ranger(
      outcome ~ ., data = train_data,
      num.trees = 500, mtry = m,
      importance = "impurity", seed = 42
    )$prediction.error
  })

  best_mtry <- mtry_candidates[which.min(cv_oob)]
  cat("Best mtry:", best_mtry, " | OOB errors:", round(cv_oob, 3), "\n")

  rf_fit <- ranger(
    outcome ~ ., data = train_data,
    num.trees = 500, mtry = best_mtry,
    importance = "impurity", seed = 42,
    probability = FALSE
  )
  saveRDS(rf_fit, rf_cache)
} else {
  rf_fit <- readRDS(rf_cache)
}

cat("OOB error rate:", round(rf_fit$prediction.error, 3), "\n")
```

```{r rf-eval}
rf_pred <- predict(rf_fit, data = test_data)$predictions
rf_cm   <- confusionMatrix(rf_pred, test_data$outcome)
cat("Random Forest — Test Accuracy:", round(rf_cm$overall["Accuracy"], 3), "\n")
print(rf_cm$table)
```

```{r rf-varimp, fig.width=8, fig.height=5}
vimp <- sort(rf_fit$variable.importance, decreasing = TRUE)

data.frame(predictor = names(vimp), importance = vimp) |>
  ggplot(aes(x = reorder(predictor, importance), y = importance)) +
  geom_col(fill = "#4575b4", alpha = 0.85) +
  coord_flip() +
  labs(title = "Random Forest Variable Importance",
       subtitle = "Mean decrease in Gini impurity",
       x = NULL, y = "Importance") +
  theme_minimal(base_size = 13)
```

---

### Model 3: Bayesian Multinomial Logistic Regression

```{r stan-model, results="hide", message=FALSE, warning=FALSE}
# Reduced predictor set for faster sampling
bayes_vars <- c("outcome", "draft_pick", "avg_ppg", "avg_rpg",
                "avg_apg", "avg_min_pg", "avg_fg_pct")

train_b <- train_data[, bayes_vars]
test_b  <- test_data[,  bayes_vars]

# Design matrices (no intercept column; Stan adds it)
X_train <- model.matrix(outcome ~ . - 1, data = train_b)
X_test  <- model.matrix(outcome ~ . - 1, data = test_b)

# Scale predictors
X_scale <- scale(X_train)
X_train_s <- as.matrix(X_scale)
X_test_s  <- scale(X_test,
                   center = attr(X_scale, "scaled:center"),
                   scale  = attr(X_scale, "scaled:scale"))

K <- nlevels(train_b$outcome)   # 4 classes
y_train <- as.integer(train_b$outcome)

stan_cache <- "models/bayes_fit.rds"
if (!file.exists(stan_cache)) {
  stan_data <- list(
    N      = nrow(X_train_s),
    K      = K,
    P      = ncol(X_train_s),
    X      = X_train_s,
    y      = y_train,
    N_test = nrow(X_test_s),
    X_test = X_test_s
  )

  stan_code <- "
data {
  int<lower=1> N;
  int<lower=2> K;
  int<lower=1> P;
  matrix[N, P] X;
  array[N] int<lower=1, upper=K> y;
  int<lower=1> N_test;
  matrix[N_test, P] X_test;
}
parameters {
  matrix[K-1, P] beta;       // coefficients (K-1 classes vs reference)
  vector[K-1]    alpha;      // intercepts
}
model {
  to_vector(beta) ~ normal(0, 2.5);
  alpha           ~ normal(0, 2.5);
  {
    matrix[N, K] log_lik;
    log_lik[, 1] = rep_vector(0, N);
    for (k in 2:K)
      log_lik[, k] = alpha[k-1] + X * beta[k-1]';
    for (n in 1:N)
      target += log_softmax(log_lik[n]')[y[n]];
  }
}
generated quantities {
  // Posterior class probabilities for test observations
  matrix[N_test, K] prob_test;
  {
    matrix[N_test, K] log_lik_test;
    log_lik_test[, 1] = rep_vector(0, N_test);
    for (k in 2:K)
      log_lik_test[, k] = alpha[k-1] + X_test * beta[k-1]';
    for (n in 1:N_test)
      prob_test[n] = softmax(log_lik_test[n]')';
  }
}
"

  bayes_fit <- stan(
    model_code = stan_code,
    data       = stan_data,
    chains     = 4,
    iter       = 2000,
    warmup     = 1000,
    seed       = 42,
    refresh    = 0
  )
  saveRDS(bayes_fit, stan_cache)
} else {
  bayes_fit <- readRDS(stan_cache)
}
```

```{r bayes-diagnostics}
# R-hat check
rhat_vals <- summary(bayes_fit)$summary[, "Rhat"]
cat("Max R-hat:", round(max(rhat_vals, na.rm = TRUE), 3), "\n")
cat("Parameters with R-hat > 1.05:", sum(rhat_vals > 1.05, na.rm = TRUE), "\n")
```

```{r bayes-coef-plot, fig.width=9, fig.height=5}
outcome_levels <- levels(train_b$outcome)
pred_names     <- colnames(X_train_s)

# Extract beta draws and plot 90% credible intervals
posterior_draws <- as.array(bayes_fit, pars = "beta")

# Rename dimensions for bayesplot
dimnames(posterior_draws)[[3]] <- paste0(
  "beta[",
  rep(outcome_levels[-1], each = length(pred_names)),
  ", ", rep(pred_names, times = K - 1), "]"
)

mcmc_intervals(posterior_draws, prob = 0.9, prob_outer = 0.95) +
  labs(title = "Bayesian MLR: 90% Posterior Credible Intervals",
       subtitle = "Coefficients for each outcome class vs. Out of League baseline") +
  theme_minimal(base_size = 12)
```

```{r bayes-pred}
# Posterior mean class probabilities on test set → point prediction
prob_test_draws <- extract(bayes_fit, pars = "prob_test")$prob_test
prob_test_mean  <- apply(prob_test_draws, c(2, 3), mean)
bayes_pred_idx  <- apply(prob_test_mean, 1, which.max)
bayes_pred      <- factor(outcome_levels[bayes_pred_idx], levels = outcome_levels)

bayes_cm <- confusionMatrix(bayes_pred, test_b$outcome)
cat("Bayesian MLR — Test Accuracy:", round(bayes_cm$overall["Accuracy"], 3), "\n")
print(bayes_cm$table)
```

```{r bayes-profiles, fig.width=9, fig.height=4}
# Posterior class probabilities for two illustrative player profiles:
# Profile A: lottery pick (pick 5), high early production
# Profile B: late second round (pick 55), modest production
profiles <- data.frame(
  draft_pick  = c(5,   55),
  avg_ppg     = c(16,   7),
  avg_rpg     = c(5,    3),
  avg_apg     = c(4,    2),
  avg_min_pg  = c(30,  15),
  avg_fg_pct  = c(0.46, 0.42)
)

profiles_s <- scale(profiles,
                    center = attr(X_scale, "scaled:center"),
                    scale  = attr(X_scale, "scaled:scale"))

# Re-use posterior draws of alpha and beta to get profile probabilities
alpha_draws <- extract(bayes_fit, pars = "alpha")$alpha   # [S, K-1]
beta_draws  <- extract(bayes_fit, pars = "beta")$beta      # [S, K-1, P]
S <- nrow(alpha_draws)

profile_probs <- lapply(1:nrow(profiles_s), function(i) {
  x_i <- as.numeric(profiles_s[i, ])
  # beta_draws is [S, K-1, P]; compute linear predictor for each class per draw
  lin_pred <- matrix(0, nrow = S, ncol = K - 1)
  for (k in seq_len(K - 1)) {
    lin_pred[, k] <- alpha_draws[, k] + beta_draws[, k, ] %*% x_i
  }
  log_lik_mat <- cbind(0, lin_pred)   # [S, K], reference class score = 0
  probs <- t(apply(log_lik_mat, 1, function(row) {
    e <- exp(row - max(row))           # numerically stable softmax
    e / sum(e)
  }))
  data.frame(
    profile = ifelse(i == 1, "Lottery Pick (5th, ~16 PPG)", "Late 2nd Round (55th, ~7 PPG)"),
    outcome = outcome_levels,
    mean    = colMeans(probs),
    lower   = apply(probs, 2, quantile, 0.05),
    upper   = apply(probs, 2, quantile, 0.95)
  )
}) |> bind_rows()

profile_probs |>
  mutate(outcome = factor(outcome, levels = outcome_levels)) |>
  ggplot(aes(x = outcome, y = mean, fill = outcome)) +
  geom_col(alpha = 0.8) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.25) +
  facet_wrap(~profile) +
  scale_fill_manual(values = c(
    "Out of League" = "#d73027", "Role Player" = "#fc8d59",
    "Starter" = "#4575b4", "All-Star" = "#1a9850"
  )) +
  labs(title = "Posterior Class Probabilities for Illustrative Player Profiles",
       subtitle = "Bars = posterior mean; error bars = 90% credible interval",
       x = NULL, y = "Probability") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 20, hjust = 1))
```

---

### Model 4: Ordinal Logistic Regression

The outcome tiers are ordered (Out of League < Role Player < Starter < All-Star), so ordinal logistic regression is a natural complement — it exploits that structure for more efficient estimates and tests whether a single latent continuum explains the outcomes.

```{r polr-fit}
polr_cache <- "models/polr_fit.rds"
if (!file.exists(polr_cache)) {
  polr_fit <- polr(
    outcome ~ draft_pick + avg_ppg + avg_rpg + avg_apg + avg_spg + avg_bpg +
      avg_fg_pct + avg_fg3_pct + avg_ft_pct + avg_min_pg + n_seasons,
    data   = train_data,
    method = "logistic",
    Hess   = TRUE
  )
  saveRDS(polr_fit, polr_cache)
} else {
  polr_fit <- readRDS(polr_cache)
}
summary(polr_fit)
```

```{r polr-ci}
polr_ci <- confint(polr_fit)   # profile-likelihood 95% CIs
kable(round(cbind(coef(polr_fit), polr_ci), 3),
      col.names = c("Estimate", "2.5%", "97.5%"),
      caption   = "Ordinal LR Coefficients with 95% Profile-Likelihood CIs")
```

```{r polr-eval}
polr_pred <- predict(polr_fit, newdata = test_data)
polr_cm   <- confusionMatrix(polr_pred, test_data$outcome)
cat("Ordinal LR — Test Accuracy:", round(polr_cm$overall["Accuracy"], 3), "\n")
print(polr_cm$table)
```

---

### Model Comparison


```{r model-comparison}
# Helper: macro-averaged F1
macro_f1 <- function(cm) {
  per_class <- cm$byClass[, "F1"]
  mean(per_class, na.rm = TRUE)
}

comparison <- data.frame(
  Model    = c("Multinomial LR", "Random Forest",
               "Bayesian MLR", "Ordinal LR"),
  Accuracy = c(
    round(mlr_cm$overall["Accuracy"],   3),
    round(rf_cm$overall["Accuracy"],    3),
    round(bayes_cm$overall["Accuracy"], 3),
    round(polr_cm$overall["Accuracy"],  3)
  ),
  Macro_F1 = c(
    round(macro_f1(mlr_cm),   3),
    round(macro_f1(rf_cm),    3),
    round(macro_f1(bayes_cm), 3),
    round(macro_f1(polr_cm),  3)
  )
)

kable(comparison, caption = "Test-set performance across all models")
```