---
title: "NBA Career Outcomes Prediction"
format: html
---

```{r setup, message=FALSE, warning=FALSE}
library(hoopR)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
```

## Project Overview

Given a player's first few seasons in the NBA, predict their career outcome tier:

- **Out of League**: washed out quickly, minimal career MPG
- **Role Player**: bench contributor, 10–20 MPG
- **Starter**: regular starter, 20+ MPG sustained
- **All-Star**: ≥1 All-Star selection

We will use multinomial logistic regression and random forest models with features derived from players' first 1–3 seasons.

---

## Data Collection

Pull league-wide per-season stats using `hoopR::nba_leaguedashplayerstats()` for seasons 2000–2015. Cache to disk to avoid re-scraping.

```{r load-season-stats, message=FALSE, warning=FALSE}
seasons <- 2000:2015
cache_file <- "data/season_stats.rds"

if (!file.exists(cache_file)) {
  dir.create("data", showWarnings = FALSE)
  season_stats_raw <- map_dfr(seasons, function(yr) {
    Sys.sleep(0.6)  # be polite to the API
    tryCatch(
      nba_leaguedashplayerstats(season = year_to_season(yr)) |>
        pluck("LeagueDashPlayerStats") |>
        mutate(season = yr),
      error = function(e) NULL
    )
  })
  saveRDS(season_stats_raw, cache_file)
} else {
  season_stats_raw <- readRDS(cache_file)
}
```

```{r clean-stats}
season_stats <- season_stats_raw |>
  transmute(
    player_id  = as.integer(PLAYER_ID),
    player     = PLAYER_NAME,
    season     = season,
    gp         = as.numeric(GP),
    min_pg     = as.numeric(MIN),
    ppg        = as.numeric(PTS),
    rpg        = as.numeric(REB),
    apg        = as.numeric(AST),
    spg        = as.numeric(STL),
    bpg        = as.numeric(BLK),
    fg_pct     = as.numeric(FG_PCT),
    fg3_pct    = as.numeric(FG3_PCT),
    ft_pct     = as.numeric(FT_PCT),
    age        = as.numeric(AGE)
  ) |>
  filter(gp >= 10)  # drop injury-shortened stints
```

## Career Outcome Labels

Label each player by their peak career tier. Pull each player's full career stats once, then classify.

```{r career-labels, message=FALSE, warning=FALSE}
all_star_cache <- "data/allstar_playerids.rds"

if (!file.exists(all_star_cache)) {
  # All-Star rosters are available via nba_allstarballotpredictor or
  # we approximate using seasons where a player averaged 20+ PPG and 3+ ASG appearances.
  # Use nba_leaguedashplayerstats filtered to All-Star seasons (2000-2015) as a proxy:
  # Players who appeared in the top scoring tier consistently are labeled All-Star.
  # For a cleaner label, pull each player's career accolades via nba_playercareerstats.
  player_ids <- unique(season_stats$player_id)

  career_raw <- map(player_ids, function(pid) {
    Sys.sleep(0.4)
    tryCatch({
      res <- nba_playercareerstats(player_id = pid)
      list(
        totals  = res |> pluck("CareerTotalsRegularSeason"),
        seasons = res |> pluck("SeasonTotalsRegularSeason")
      )
    }, error = function(e) NULL)
  })

  saveRDS(career_raw, all_star_cache)
} else {
  career_raw <- readRDS(all_star_cache)
}

# Cache was saved as a flat list of CareerTotalsRegularSeason data frames
# with a lowercase player_id column already attached
career_totals <- career_raw |>
  compact() |>
  bind_rows() |>
  transmute(
    player_id  = as.integer(player_id),
    career_gp  = as.numeric(GP),
    career_min = as.numeric(MIN),
    career_ppg = as.numeric(PTS) / as.numeric(GP),
    career_mpg = as.numeric(MIN) / as.numeric(GP),
    # Approximate seasons from observed season stats (no season-by-season data in cache)
    career_seasons = career_gp / 55
  )

# All-Star approximation: players with career PPG > 18 and 10+ seasons → All-Star
# Starter: career MPG >= 28 | PPG >= 14
# Role Player: career MPG 15–28 and career_seasons >= 4
# Out of League: everything else (short careers, low minutes)
player_labels <- career_totals |>
  mutate(
    outcome = case_when(
      career_ppg >= 18 & career_seasons >= 8             ~ "All-Star",
      career_mpg >= 28 | (career_ppg >= 14 & career_seasons >= 6) ~ "Starter",
      career_mpg >= 15 & career_seasons >= 4              ~ "Role Player",
      TRUE                                                 ~ "Out of League"
    ),
    outcome = factor(outcome,
                     levels = c("Out of League", "Role Player", "Starter", "All-Star"))
  ) |>
  select(player_id, outcome, career_ppg, career_mpg, career_seasons)
```

```{r join-draft}
# Pull draft info for each player
draft_cache <- "data/draft_info.rds"

if (!file.exists(draft_cache)) {
  draft_raw <- map(unique(season_stats$player_id), function(pid) {
    Sys.sleep(0.003)
    tryCatch(
      nba_commonplayerinfo(player_id = pid) |>
        pluck("CommonPlayerInfo") |>
        mutate(player_id = pid),
      error = function(e) NULL
    )
  })
  saveRDS(draft_raw, draft_cache)
} else {
  draft_raw <- readRDS(draft_cache)
}

draft_info <- draft_raw |>
  compact() |>
  bind_rows() |>
  transmute(
    player_id  = as.integer(player_id),
    draft_pick = suppressWarnings(as.numeric(DRAFT_NUMBER))
  ) |>
  filter(!is.na(draft_pick))
```

```{r build-model-data}
# Identify each player's debut season, then keep their first 3 seasons
debut_season <- season_stats |>
  group_by(player_id) |>
  summarise(debut = min(season), .groups = "drop")

first3 <- season_stats |>
  left_join(debut_season, by = "player_id") |>
  filter(season <= debut + 2) |>
  group_by(player_id) |>
  summarise(
    across(c(ppg, rpg, apg, spg, bpg, fg_pct, fg3_pct, ft_pct, min_pg),
           mean, na.rm = TRUE, .names = "avg_{.col}"),
    n_seasons = n(),
    .groups = "drop"
  )

model_data <- first3 |>
  inner_join(player_labels, by = "player_id") |>
  inner_join(draft_info,   by = "player_id") |>
  filter(n_seasons >= 1)

glimpse(model_data)
```

---

## Exploratory Visualizations

### Visualization 1: Draft Position by Career Outcome

```{r viz1-draft, fig.width=8, fig.height=5}
model_data |>
  filter(draft_pick <= 60) |>
  ggplot(aes(x = outcome, y = draft_pick, fill = outcome)) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.8) +
  geom_jitter(width = 0.12, size = 1.5, alpha = 0.4, color = "gray30") +
  scale_y_reverse(breaks = c(1, 10, 20, 30, 45, 60)) +
  scale_fill_manual(values = c(
    "Out of League" = "#d73027",
    "Role Player"   = "#fc8d59",
    "Starter"       = "#4575b4",
    "All-Star"      = "#1a9850"
  )) +
  labs(
    title    = "Draft Position by Career Outcome",
    subtitle = "Lower pick number = higher draft position  |  2000–2015 draft classes",
    x = "Career Outcome", y = "Draft Pick Number"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

### Visualization 2: PPG in First 3 Seasons by Career Outcome

```{r viz2-ppg, fig.width=8, fig.height=5}
model_data |>
  ggplot(aes(x = avg_ppg, fill = outcome, color = outcome)) +
  geom_density(alpha = 0.35, linewidth = 0.8) +
  scale_fill_manual(values = c(
    "Out of League" = "#d73027",
    "Role Player"   = "#fc8d59",
    "Starter"       = "#4575b4",
    "All-Star"      = "#1a9850"
  )) +
  scale_color_manual(values = c(
    "Out of League" = "#d73027",
    "Role Player"   = "#fc8d59",
    "Starter"       = "#4575b4",
    "All-Star"      = "#1a9850"
  )) +
  labs(
    title    = "PPG Distribution (First 3 Seasons) by Career Outcome",
    subtitle = "Average over each player's first 1–3 seasons",
    x = "Points Per Game", y = "Density",
    fill = "Outcome", color = "Outcome"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

### Visualization 3: Career Scoring Trajectory by Outcome Tier

```{r viz3-trajectories, fig.width=9, fig.height=5, message=FALSE, warning=FALSE}
# For a random sample of players per tier, show their season PPG over first 3 years
set.seed(42)
sampled_ids <- model_data |>
  group_by(outcome) |>
  slice_sample(n = 10) |>
  pull(player_id)

traj_data <- season_stats |>
  inner_join(debut_season, by = "player_id") |>
  filter(player_id %in% sampled_ids, season <= debut + 2) |>
  mutate(year_in_league = season - debut + 1) |>
  inner_join(select(player_labels, player_id, outcome), by = "player_id")

ggplot(traj_data, aes(x = year_in_league, y = ppg,
                       group = player_id, color = outcome)) +
  geom_line(alpha = 0.4, linewidth = 0.8) +
  geom_point(alpha = 0.6, size = 2) +
  stat_summary(aes(group = outcome), fun = mean,
               geom = "line", linewidth = 2, linetype = "dashed") +
  scale_color_manual(values = c(
    "Out of League" = "#d73027",
    "Role Player"   = "#fc8d59",
    "Starter"       = "#4575b4",
    "All-Star"      = "#1a9850"
  )) +
  scale_x_continuous(breaks = 1:3, labels = c("Year 1", "Year 2", "Year 3")) +
  labs(
    title    = "Scoring Trajectory in First 3 Seasons by Career Outcome",
    subtitle = "Thin = individual players (sample of 10/tier)  |  Dashed = tier average",
    x = "Season in League", y = "Points Per Game",
    color = "Outcome"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "right")
```
